---
title: "eegSeminaR Notes"
author: "Daniel Soto"
---

These are my notes and completed exercises taken from [this 2022 seminar on EEG data analysis in R](https://alexenge.github.io/eegSeminaR/). 
Citation:
> Enge A (2023). eegSeminaR: A Seminar on Analyzing EEG Data in R. https://github.com/alexenge/eegSeminaR, https://alexenge.github.io/eegSeminaR/.


## Setting Up Data

Though they use the data at `alexenge/erpcore`, we will be downloading the data using the `osfr` package, as the `erpcore` R package seems to be broken.

```{r}
#| label: install-osfr
#| output: false

pak::pak("osfr")

library(osfr)
library(here)    # For general handling of local data files
```

Moreover, we will only be using data from the N170 experiment, which showed that subjects who were hooked up to EEGs and were tasked with viewing faces elicited different responses than when they were tasked to view cars.
We use the `osfr` package to get the data in BIDS format.

```{r}
#| label: get-n170
#| eval: false

bids_dir <- here("eegs", "data/n170")
dir.create(bids_dir, recursive = TRUE)

# Connect to the N170 OSF project
n170_osf <- osf_retrieve_node("https://osf.io/pfde9/")

# See the top-level folders
folders <- osf_ls_files(n170_osf, n_max = Inf)
folders

# Get the folder with data in BIDS format
bids_folder <- folders[
  folders$name == "N170 Raw Data BIDS-Compatible",
]
bids_folder

bids_files <- osf_ls_files(
  bids_folder,
  n_max = Inf
)

osf_download(
  bids_files,
  path = bids_dir,
  conflicts = "overwrite",
  recurse = TRUE
)
```

## Preprocessing

For preprocessing, we first load `eegUtils`, a package for working with EEG data.

```{r}
#| label: load-eegUtils
#| output: false

pak::pak("craddm/eegUtils")

library(eegUtils)
```

The downloaded ERP CORE data is in `data/n170`, as we specified in the setup section.
The data is in BIDS format; though the seminar notes say the subjects are each in their respective `sub-XXX/egg` directory, on my end the file structure looks like `{subject-number}/{subject-number}_N170.set`.

So, we retrieve this `.set` file for subject 1:

```{r}
#| label: set-file

bids_dir <- here("eegs", "data/n170")
set_file <- here(bids_dir, "sub-001/eeg/sub-001_task-N170_eeg.set")
file.exists(set_file)
```
`eegUtils` has functions for importing raw EEG data from various file formats into R, including the `.set` format used by ERP CORE.

```{r}
#| label: import-set-deps
#| output: false

# Needed these packages for `import_set` to work
pak::pak("R.matlab", "hdf5r") 
```

```{r}
#| label: data-raw

dat_raw <- import_set(set_file)
```

Inspecting the data type:
```{r}
#| label: inspect-data-type

class(dat_raw)
```

we see `eeg_data` is a custom class defined by `eegUtils` — a big list with many sub-components:

```{r}
#| label: inspect-dat-raw

str(dat_raw)
```

The data we actually care about is in the `dat_raw$signals` data frame.
Here, each row is one sample (time point), each column is one channel (EEG electrode), and each value is the EEG voltage (in microvolts; µV) measured at this time sample and channel.

### Viewing Data

We can visually check the data at multiple steps using the `browse_data` function. 
This will open a site we can open either on our browser or in the "Viewer" tab on the right-hand side of RStudio from which we can interactively inspect the data.

```{r}
#| label: browse-data
#| eval: false
#| output: false

browse_data(dat_raw)
```

We can also make a static butterfly plot of a given chunk of data:

```{r}
#| label: butterfly-plot

time_lim <- c(300, 305)
plot_butterfly(dat_raw, time_lim, baseline = time_lim)
```

### Channel Locations

For some steps, such as with making topographic plots, we need to know the relative positions of electrodes on the scalp. 
These are sometimes included in the raw data, but in this case we need to use a separate function:

```{r}
#| label: electrode-locations

dat_raw <- electrode_locations(dat_raw, montage = "biosemi64", overwrite = TRUE)
dat_raw$chan_info

plot_electrodes(dat_raw, interact = TRUE)
```

### Re-referencing

During recording ("online"), each electrode is measuring the difference in voltage compared to the reference electrode (e.g., FCz).
In contrast, during preprocessing ("offline"), a new reference is typically chose:

- **Average reference**: Difference to the average of all scalp channels
- **Linked mastoids**: Difference to the average of the mastoid channels (M1 and M2)

```{r}
#| label: re-referencing

eog_channels <- c("HEOG_left", "HEOG_right", "VEOG_lower")
dat_reref <- eeg_reference(dat_raw, ref_chans = "average", exclude = eog_channels)
```

### Filtering

We now want to filter our signals for frequencies that represent brain waves, and nothing else.
According to the seminar's notes, the kinds of wave frequencies that we will encounter will represent the following EEG activity:

Frequency	| EEG activity
--- | ---
< 0.1 Hz |	Electrode drift, sweat artifacts
0.1–4 Hz (delta) |	Slow-wave sleep
4–8 Hz (theta) |	Drowsiness, idling
8–13 Hz (alpha) |	Relaxing, inhibition
13–30 Hz (beta)	| Active thinking, alertness
30–80 Hz (gamma) | Binding, consciousness (?)
50 Hz (US: 60 Hz)	| Power line noise
\> 80 Hz	| Muscle artifacts

Thus, we only really care about the waves with frequencies between 0.1 Hz and 30 Hz.
To remove unwanted parts of the signal, we do the following:

1. Convert time-domain EEG into the frequency domain via Fourier Transform
2. Remove or attenuate certain frequencies, depending on whether they are in our range or not.
  - A *high-pass* filter will remove slow frequencies (in our case, <0.1 Hz)
  - A *low-pass* filter will remove fast frequencies (in our case, >30 Hz)
  - A *band-pass* filter combines a high-pass and low-pass filter
  - A *band-stop* filter removes a narrow frequency range (e.g. around 50 Hz)
3. Convert cleaned signal back into the time domain

This can all be done with one function, `eeg_filter`:

```{r}
#| label: filtering

dat_filt <- eeg_filter(dat_reref, low_freq = 0.1, high_freq = 30.0)
plot_butterfly(dat_filt, time_lim, baseline = time_lim)
```

Let's now save our preprocessed data: 

```{r}
#| label: save-preprocessed-data
#| output: false

deriv_dir <- here(bids_dir, "derivatives/eegUtils/sub-001/eeg")
preproc_file <- here(deriv_dir, "sub-001_task-N170_desc-preprocessed_eeg.rds")
dir.create(deriv_dir, recursive = TRUE)
saveRDS(dat_filt, preproc_file)
```

### Extra: Downsampling

The **sampling rate** is the number of EEG voltages recorded per second (in Hz).
```{r}
#| label: sampling-rate

dat_raw$srate
```

This means that each pair of samples is less than 1 ms apart:
```{r}
#| label: samp-rate-recip

1 / dat_raw$srate
```
We don't need that much temporal resolution, so we will downsample the data using `eeg_downsample` from the `eegUtils` library:

```{r}
#| label: downsampling

dat_downs <- eeg_downsample(dat_filt, q = 4.0) # q is the downsampling factor
```

This will reduce our computer's memory load and make subsequent computations faster:

```{r}
#| label: memory-load

show_size_mb <- function(x) format(object.size(x), units = "MB")
lapply(list(before = dat_filt, after = dat_downs), show_size_mb)
saveRDS(dat_downs, preproc_file)
```

## Creating Epochs

```{r}
#| label: set-deriv-dir

deriv_dir <- here(bids_dir, "derivatives/eegUtils/sub-001/eeg")
```

### Event Codes

Significant "events," (e.g. stimuli, responses) are marked in the EEG recording by unique **event codes**.
In this experiment, event codes between 1-40 mean the stimulus was a face, while event codes between 41-80 mean the stimulus was a car.
Note that in my case, the `task-N170_events.json` file did not exist in the directory where the N170 data got downloaded, but that is ok because we still know from the paper that this convention was used.
We can hard-code these events as epoch labels as follows:

```{r}
#| label: event-codes

event_codes <- c(1:80)
epoch_labels <- ifelse(event_codes <= 40, "face", "car")
```

### Construct Epochs

Now, knowing event codes, we can "cut out" our epochs around them.
For the epoch data, we want to include between 100-200 ms before the stimulus onset as a baseline period.

```{r}
#| label: epoch-data

dat_epo <- epoch_data(
  dat_downs,
  event_codes,
  time_lim = c(-0.2, 0.8),
  epoch_labels = epoch_labels
)
```

To plot these epochs, we use an "ERP image," with:

- X-axis: Time relative to stimulus onset
- Y-axis: Trial number (or any other sorting, e.g. by reaction time)
- Colors: EEG amplitude

```{r}
#| label: erp-image

erp_image(dat_epo, electrode = "PO8", smoothing = 1)
```

### Baseline Correction

Trials will differ in their overall amplitude, i.e. due to drifts and slow artifacts. 
This can hide more subtle differences between conditions.
The solution to this is to subtract the average activity during the pre-stimulus baseline:

```{r}
#| label: remove-baseline

dat_basl <- rm_baseline(dat_epo, c(-0.2, 0.0))
erp_image(dat_basl, electrode = "PO4", smoothing = 1)
```

We can confirm that the average voltage during the baseline window is actually zero given an epoch and channel:

```{r}
#| label: voltage-confirmation

# Extract baseline time points (baseline meaning before onset, thus t < 0.0) at 
# epoch 42
inds <- which(dat_basl$timings$epoch == 42 & dat_basl$timings$time < 0.0)

mean(dat_basl$signals$FP1[inds]) # close enough to 0
```
Now, we can save the baseline-corrected epochs.

```{r}
#| label: save-epochs

epo_file <- here(deriv_dir, "sub-001_task-N170_desc-epoched_eeg.rds")
saveRDS(dat_basl, epo_file)
```
## Removing Artifacts

### Independent Component Analysis

With the baseline-corrected epoch data, we can run Independent Component Analysis (ICA) to separate our EEG recordings into a number of additive sub-components.
The seminar decides to separate into the 15 most important components:

```{r}
#| label: run-ica

n_components <- 15
ica <- run_ICA(dat_epo, method = "fastica", pca = n_components)
```

With this, we can plot scalp distributions of the extracted ICA components. 
In particular, eye-blink and vertical eye movement components will have a very strong signal at the forehead (`Comp001``Comp001`), and vertical eye movement components will have opposite polarities left-frontal vs. right-frontal (maybe `Comp008` and/or `Comp009`).

```{r}
#| label: ica-topoplot

topoplot(
  ica,
  component = 1:n_components,
  limits = c(-20, 20),
  chan_marker = "none",
  scaling=0.4
)
```

We can also automatically detect "bad" components based on their correlation with the electrooculogram (EOG), the part that measures eye-movement signals.
We set a threshold of `0.6`:

```{r}
#| label: eog-corr

heog <- c("HEOG_left", "HEOG_right")
veog <- c("FP2", "VEOG_lower")
bad_components <- ar_eogcor(ica, dat_epo, heog, veog, threshold = 0.6)
bad_components
```

Weirdly, in the seminar, it ends up being components 001 and 009 that are outliers, but here we see we get components 001 and 006 are the outliers.
I chalk this up to the potential differences that may have come from us using different R package sources for the data we use.

Regardless, with this, we can remove the "bad" ICA components from the actual (epoched) data:

```{r}
#| label: remove-bad-ica-comps

dat_icacorr <- apply_ica(dat_epo, ica, comps = bad_components)
```
The seminar provides useful helper functions to compare the uncorrected and corrected epochs visually:

```{r}
#| label: bad-comp-removal-visual

extract_epoch <- function(epochs, epoch_no, electrode) {
  epoch_ixs <- epochs$timings$epoch == epoch_no
  data.frame(
    time = unique(epochs$timings$time),
    amplitude = epochs$signals[epoch_ixs, electrode][[1]]
  )
}

compare_epochs <- function(uncorr_epochs,
                           corr_epochs,
                           epoch_no,
                           electrode) {
  require(ggplot2)
  n_samples <- length(unique(uncorr_epochs$timings$time))
  epochs <- list(uncorr_epochs, corr_epochs) |>
    lapply(extract_epoch, epoch_no, electrode) |>
    do.call(what = "rbind")
  epochs$type <- rep(c("uncorrected", "corrected"), each = n_samples) |>
    factor(levels = c("uncorrected", "corrected"))
  ggplot(epochs, aes(x = time, y = amplitude, color = type)) +
    geom_line() +
    labs(x = "Time (s)", y = "Amplitude (µV)", color = NULL) +
    ggtitle(paste0("Epoch #", epoch_no, ", channel ", electrode)) +
    theme_minimal()
}

compare_epochs(dat_epo, dat_icacorr, epoch_no = 103, electrode = "FP1")
```

### Artifact Rejection

ICA may not be able to correct all eye movements or other artifacts.
Additionally, we want to reject all epochs where at least one channel exceeds an absolute voltage thresheld, like ±100 µV.

```{r}
#| label: artifact-rejection

dat_rej <- ar_thresh(dat_icacorr, threshold = c(100, -100), reject = TRUE)
```

Thresholds in practice should be chosen carefully, as rejecting many epochs reduces statistical power. 
More sophisticated thresholding approaches exist, e.g. via a moving window.
Save artifact-corrected epochs for later:

```{r}
#| label: save-artifact-corr-epoch-dat

corr_file <- here(deriv_dir, "sub-001_task-N170_desc-corrected_eeg.rds")
saveRDS(dat_rej, corr_file)
```

## Extracting Amplitudes

### Selecting Region of Interest

We can't just measure a single EEG electrode. 
For one thing, single measurements are noisy and spatially imprecise.
Thus, averaging across nearby electrodes boosts our signal-to-noise ratio.
For another, their measurements are not well-defined without relationships to other electrodes.
Altogether, this leads to us defining the **region of interest (ROI)**, "the set of electrodes where we expect our ERP component of interest to be maximal."

First, we convert our epoched data to a data frame for subsetting / averages.

```{r}
#| label: epoched-data-frame

dat_frame <- as.data.frame(dat_epo)
str(dat_frame)
```
Now, we know from prior N170 research that the face-sensitive N170 is strongest over bilateral posterior-occipitotemporal scalp, so we'll summarize the EEG from the six electrodes covering that region:

| Left Hemisphere | Right Hemisphere |
| ---- | ----- |
| P7   | P8    |
| P9   | P10   |
| PO7  | PO8   |

In particular, we have

- P7, P8 = left/right parietal
- P9, P10 = farther lateral/inferior left/right posterior sites
- PO7, PO8 = left/right parieto-occipital

With this, we create a new "virtual" channel, a column in the data frame, containing the average signal in the ROI:

```{r}
#| label: virtual-channel

roi <- c("P7", "P8", "P9", "P10", "PO7", "PO8")
dat_frame$roi <- rowMeans(dat_frame[roi])
```

### Select Time Window

ERP components unfold on the time scale of 10 ms to multiples 100s of ms.
Here, we want to select the time window for where we expect our component to be — in our case, around 170 ms for the N170.

```{r}
#| label: select-time-window

tmin <- 0.15
tmax <- 0.2

dat_frame <- subset(dat_frame, time >= tmin & time < tmax)
```

Averaging across this time window gets us one value, the mean ERP amplitude, per trial:

```{r}
#| label: averaging-time-window

dat_trials <- aggregate(
  roi ~ epoch + event_type + epoch_labels,
  dat_frame,
  mean
)
names(dat_trials) <- c("epoch", "stimulus", "condition", "amplitude")
str(dat_trials)
```
We can make a quick box plot comparing the single trial mean ERP amplitudes for faces vs. cars:

```{r}
#| label: dat-trials-boxplot

ggplot(dat_trials, aes(x = condition, y = amplitude)) +
  geom_boxplot() +
  labs(x = NULL, y = "N170 amplitude (µV)") +
  theme_minimal()
```

We’ll quantify this effect using statistical modeling in Vignette 7. 
Finally, we save the single trial data for later:

```{r}
#| label: save-trials-file

trials_file <- here(deriv_dir, "sub-001_task-N170_trials.csv")
write.csv(dat_trials, trials_file, row.names = FALSE)
```
