---
title: "EEG Seminar Notes"
author: "Daniel Soto"
---

These are my notes and completed exercises taken from the seminar notes at https://alexenge.github.io/eegSeminaR/.

## Setting Up Data

Though they use the data at `alexenge/erpcore`, we will use the data from `eegverse/erpcore`, as the former does not seem to exist anymore on GitHub.

```{r}
#| label: install-erpcore
#| output: false

pak::pak("eegverse/erpcore")

library(erpcore)
```

Moreover, we will only be using data from the N170 experiment, which showed that subjects who were hooked up to EEGs and were tasked with viewing faces elicited different responses than when they were tasked to view cars.

```{r}
#| label: get-n170
#| eval: false

library(here)

data_dir <- here("eegs", "data/n170") # for me, all of my files are in ./eegs/ 
dir.create(data_dir, recursive = TRUE)

# This function downloads all n170 data to `data_dir`
get_erpcore(
  component = "n170",
  dest_path = data_dir,
  type = "bids",          # we only want data in BIDS format
  subjects = "sub-001",   # limit our subjects  to just 001 for now
  conflicts = "overwrite"
)
```

## Preprocessing

For preprocessing, we first load `eegUtils`, a package for working with EEG data.

```{r}
#| label: load-eegUtils
#| output: false

pak::pak("craddm/eegUtils")

library(eegUtils)
```

The downloaded ERP CORE data is in `data/n170`, as we specified in the setup section.
The data is in BIDS format; though the seminar notes say the subjects are each in their respective `sub-XXX/egg` directory, on my end the file structure looks like `{subject-number}/{subject-number}_N170.set`.

So, we retrieve this `.set` file for subject 1:

```{r}
#| label: set-file

bids_dir <- here("eegs", "data/n170")
set_file <- here(bids_dir, "1/1_N170.set")
file.exists(set_file)
```
`eegUtils` has functions for importing raw EEG data from various file formats into R, including the `.set` format used by ERP CORE.

```{r}
#| label: import-set-deps
#| output: false

# Needed these packages for `import_set` to work
pak::pak("R.matlab", "hdf5r") 
```

```{r}
#| label: data-raw

dat_raw <- eegUtils::import_set(set_file)
```

Inspecting the data type:
```{r}
#| label: inspect-data-type

class(dat_raw)
```

we see `eeg_data` is a custom class defined by `eegUtils` — a big list with many sub-components:

```{r}
#| label: inspect-dat-raw

str(dat_raw)
```

The data we actually care about is in the `dat_raw$signals` data frame.
Here, each row is one sample (time point), each column is one channel (EEG electrode), and each value is the EEG voltage (in microvolts; µV) measured at this time sample and channel.
