---
title: "Data and Research Question"
editor_options: 
  chunk_output_type: console
---

## Youth Risk Behavior Survey 2023

The Youth Risk Behavior Survey (YRBS) is a national survey that monitors health-related behaviors among high school students, including weapon carrying and associated risk factors.

### Dataset Overview

-   **Source**: Centers for Disease Control and Prevention (CDC)
-   **Year**: 2023
-   **Target Population**: High school students
-   **Sample Size**: Approximately 19,000 students nationwide

### Research Question

**How do logistic regression, lasso, k-nearest neighbors, and tree-based models compare in predicting school-based weapon carrying among adolescents based on risk and protective factors?**

### Key Variables for the Research Question

The dataset includes information on various health-related behaviors:

1.  **Outcome**
    -   Weapon Carrying (Carried a weapon on school property)

2.  **Predictors**
    -   Traumatic experiences
    -   School Safety Perceptions
    -   Bullying Experiences
    -   Family Support
    -   Social Media Use
    -   Peer Relationships

### Environment Setup

```{r}
#| label: data-prep
#| output: false

library(tidyverse)
library(tidymodels)
library(dissertationData)
library(here)
data(clean_yrbs_2023)
```

### Creation of the Dataset

```{r}
#| label: create-dataset

analysis_data <- clean_yrbs_2023 %>%
    select(
        WeaponCarryingSchool, AttackedInNeighborhood, Bullying,
        SexualAbuseByOlderPerson, ParentalPhysicalAbuse, ParentSubstanceUse,
        ParentIncarceration, SchoolConnectedness, ParentalMonitoring,
        UnfairDisciplineAtSchool, Homelessness
    ) |>
    filter(!is.na(WeaponCarryingSchool)) %>%
    mutate(across(
        c(
            ParentSubstanceUse, ParentIncarceration, SchoolConnectedness,
            ParentalMonitoring, UnfairDisciplineAtSchool
        ),
        ~ as.numeric(.x) - 1
    )) %>%
    mutate(across(
        c(
            ParentSubstanceUse, ParentIncarceration, SchoolConnectedness,
            ParentalMonitoring, UnfairDisciplineAtSchool
        ),
        ~ factor(.x)
    ))



```

### Splitting the Dataset

```{r}
#| label: split-dataset

set.seed(1990)

analysis_split <- initial_split(analysis_data,
    strata = WeaponCarryingSchool
)

analysis_train <- training(analysis_split)
analysis_test <- testing(analysis_split)

analysis_split



```

### Cross-Validation

```{r}
#| label: cross-validation

analysis_folds <- vfold_cv(analysis_train,
    v = 5
)
analysis_folds


```