---
title: "Decision Trees"
author: "Daniel Soto"
---

In this note, we check whether a decision tree can predict whether a teen has ever consumed alcohol.

## Setup

Like before we load the packages necessary for the analysis.

```{r}
#| label: packages
#| output: false

library(tidyverse)
library(tidymodels)
library(MLearnYRBSS)
library(janitor)
library(skimr)
library(vip)
library(rpart.plot)

data("riskyBehaviors")
```

### Recode the outcome

We encode our outcome variable, `UseAlcohol`, so that it is 0 if a given teen has never used alcohol and 1 if they have used alcohol.

```{r}
#| label: outcome-encoding

riskyBehaviors_analysis <-
    riskyBehaviors |>
    mutate(
        UsedAlcohol = case_when(
            AgeFirstAlcohol == 1 ~ 0,
            AgeFirstAlcohol %in% c(2, 3, 4, 5, 6, 7) ~ 1,
            .default = NA
        ),
        UsedAlcohol = factor(UsedAlcohol)
    ) |>
    drop_na(UsedAlcohol) |>
    select(-c(AgeFirstAlcohol, DaysAlcohol, BingeDrinking,
              LargestNumberOfDrinks, SourceAlcohol))
```

We drop `AgeFirstAlcohol`, `DaysAlcohol`, `BingeDrinking`, `LargestNumberOfDrinks`, and `SourceAlcohol` because they are all about alcohol use. 
If we left them in, the tree would just memorize the answer.

### Split + Resample

Here, we stratify the train/test split.

```{r}
#| label: train-test-stratification

set.seed(2023)

alcohol_split <- initial_split(
    riskyBehaviors_analysis,
    strata = UsedAlcohol
)

alcohol_train <- training(alcohol_split)
alcohol_test <- testing(alcohol_split)

alcohol_split
```

We use `strata = ` to keep the proportion of `UsedAlcohol` roughly equal in the train and test sets. 
We use the same logic in the Lasso training.
To make sure this worked, we check the class balance:
```{r}
#| label: class-balance-check

alcohol_train |>
    tabyl(UsedAlcohol) |>
    adorn_pct_formatting(0) |>
    adorn_totals()

alcohol_test |>
    tabyl(UsedAlcohol) |>
    adorn_pct_formatting(0) |>
    adorn_totals()
```
Both sets show the same proportions, meaning `strata =` did its job.

### Cross-validation folds

`v = 10` folds of the training set. 
Every candidate tree we try will be evaluated on all 10 — the test set stays untouched until the very end.

```{r}
#| label: cv-folds

set.seed(2023)

cv_alcohol <- vfold_cv(alcohol_train, v = 10, strata = UsedAlcohol)
cv_alcohol
```

## Recipe (preprocessing)

Trees don't need fancy preprocessing: no `step_normalize()`, as trees only care about *order*, not scale; no `step_dummy()`, as `rpart` handles factor predictors natively; and imputation is still useful, since `rpart` can't split on `NA`. 

```{r}
#| label: recipe

alcohol_recipe <-
    recipe(UsedAlcohol ~ ., data = alcohol_train) |>
    step_impute_mode(all_nominal_predictors()) |>
    step_impute_mean(all_numeric_predictors())

alcohol_recipe
```

