---
title: "Decision Trees"
author: "Daniel Soto"
---

In this note, we check whether a decision tree can predict whether a teen has ever consumed alcohol.

## Setup

Like before we load the packages necessary for the analysis.

```{r}
#| label: packages
#| output: false

library(tidyverse)
library(tidymodels)
library(MLearnYRBSS)
library(janitor)
library(skimr)
library(vip)
library(rpart.plot)

data("riskyBehaviors")
```

### Recode the outcome

We encode our outcome variable, `UseAlcohol`, so that it is 0 if a given teen has never used alcohol and 1 if they have used alcohol.

```{r}
#| label: outcome-encoding

riskyBehaviors_analysis <-
    riskyBehaviors |>
    mutate(
        UsedAlcohol = case_when(
            AgeFirstAlcohol == 1 ~ 0,
            AgeFirstAlcohol %in% c(2, 3, 4, 5, 6, 7) ~ 1,
            .default = NA
        ),
        UsedAlcohol = factor(UsedAlcohol)
    ) |>
    drop_na(UsedAlcohol) |>
    select(-c(AgeFirstAlcohol, DaysAlcohol, BingeDrinking,
              LargestNumberOfDrinks, SourceAlcohol))
```

We drop `AgeFirstAlcohol`, `DaysAlcohol`, `BingeDrinking`, `LargestNumberOfDrinks`, and `SourceAlcohol` because they are all about alcohol use. 
If we left them in, the tree would just memorize the answer.

### Split + Resample

Here, we stratify the train/test split.

```{r}
#| label: train-test-stratification

set.seed(2023)

alcohol_split <- initial_split(
    riskyBehaviors_analysis,
    strata = UsedAlcohol
)

alcohol_train <- training(alcohol_split)
alcohol_test <- testing(alcohol_split)

alcohol_split
```

We use `strata = ` to keep the proportion of `UsedAlcohol` roughly equal in the train and test sets. 
We use the same logic in the Lasso training.
To make sure this worked, we check the class balance:
```{r}
#| label: class-balance-check

alcohol_train |>
    tabyl(UsedAlcohol) |>
    adorn_pct_formatting(0) |>
    adorn_totals()

alcohol_test |>
    tabyl(UsedAlcohol) |>
    adorn_pct_formatting(0) |>
    adorn_totals()
```
Both sets show the same proportions, meaning `strata =` did its job.

### Cross-validation folds

`v = 10` folds of the training set. 
Every candidate tree we try will be evaluated on all 10 — the test set stays untouched until the very end.

```{r}
#| label: cv-folds

set.seed(2023)

cv_alcohol <- vfold_cv(alcohol_train, v = 10, strata = UsedAlcohol)
cv_alcohol
```

## Workflow

### Recipe (preprocessing)

Trees don't need fancy preprocessing: no `step_normalize()`, as trees only care about *order*, not scale; no `step_dummy()`, as `rpart` handles factor predictors natively; and imputation is still useful, since `rpart` can't split on `NA`. 

```{r}
#| label: recipe

alcohol_recipe <-
    recipe(UsedAlcohol ~ ., data = alcohol_train) |>
    step_impute_mode(all_nominal_predictors()) |>
    step_impute_mean(all_numeric_predictors())

alcohol_recipe
```

### Specification

```{r}
#| label: cart-spec

cart_spec <-
    decision_tree(
        cost_complexity = tune(), # ⬅️ tune
        tree_depth = tune(),      # ⬅️ tune
        min_n = tune()            # ⬅️ tune
    ) |>
    set_engine("rpart") |>
    set_mode("classification")

cart_spec
```

### Workflow

Bundle it! This is the recipe and specification, neatly packaged.

```{r}
#| label: workflow

cart_workflow <-
    workflow() |>
    add_recipe(alcohol_recipe) |>
    add_model(cart_spec)

cart_workflow
```

## Tree tuning

### Tuning grid

We first set up the tuning grid. 

```{r}
#| label: tuning-grid

tree_grid <- grid_regular(
    cost_complexity(),
    tree_depth(c(2, 5)),
    min_n(),
    levels = 4
)

tree_grid
```

### Run the tuning

Note for those following along that this part can be slow:

- 4 × 4 × 4 = **64 candidates**
- × 10 folds = **640 model fits**
- We’ll use `doParallel` to speed it up.

`levels` can be set to `3` for faster turnaround.
We use ROC AUC as the tuning metric — robust to class imbalance, and the same one we used for Lasso. 

```{r}
#| label: run-tuning

doParallel::registerDoParallel()

cart_tune <-
    cart_workflow |>
    tune_grid(
        resamples = cv_alcohol,
        grid = tree_grid,
        metrics = metric_set(roc_auc),
        control = control_grid(save_pred = TRUE)
    )

doParallel::stopImplicitCluster()
```

### Show the best candidates

The top 5 hyperparameter combos, ranked by mean ROC AUC across the 10 folds.

```{r}
show_best(cart_tune, metric = "roc_auc")
```

### Visualize the tuning results

In this visualization, each panel shows how the metric changes as one knob varies — colored by the others. 
We look for the sweet spot.

```{r}
autoplot(cart_tune)
```

### Picking the best combination

We select one row as the winner candidate:

```{r}
best_cart <- select_best(cart_tune, metric = "roc_auc")

best_cart
```

But we don't commit yet: we want to check how this performs first.

## How does our chosen candidate actually do?

Here, we get an honest, CV-estimated performance for our single chosen combo.

```{r}
cart_workflow |>
    finalize_workflow(best_cart) |>
    fit_resamples(resamples = cv_alcohol) |>
    collect_metrics()
```

Looks good, time to lock it in.

```{r}
cart_final_wf <- finalize_workflow(cart_workflow, best_cart)

cart_final_wf
```

finalize_workflow() swaps the tune() placeholders for the actual best values. 
Now we’re committed.

