---
title: "Decision Trees"
author: "Daniel Soto"
---

In this note, we check whether a decision tree can predict whether a teen has ever consumed alcohol.

## Setup

Like before we load the packages necessary for the analysis.

```{r}
#| label: packages
#| output: false

library(tidyverse)
library(tidymodels)
library(MLearnYRBSS)
library(janitor)
library(skimr)
library(vip)
library(rpart.plot)

data("riskyBehaviors")
```

## Recode the outcome

We encode our outcome variable, `UseAlcohol`, so that it is 0 if a given teen has never used alcohol and 1 if they have used alcohol.

```{r}
#| label: outcome-encoding

riskyBehaviors_analysis <-
    riskyBehaviors |>
    mutate(
        UsedAlcohol = case_when(
            AgeFirstAlcohol == 1 ~ 0,
            AgeFirstAlcohol %in% c(2, 3, 4, 5, 6, 7) ~ 1,
            .default = NA
        ),
        UsedAlcohol = factor(UsedAlcohol)
    ) |>
    drop_na(UsedAlcohol) |>
    select(-c(AgeFirstAlcohol, DaysAlcohol, BingeDrinking,
              LargestNumberOfDrinks, SourceAlcohol))
```

We drop `AgeFirstAlcohol`, `DaysAlcohol`, `BingeDrinking`, `LargestNumberOfDrinks`, and `SourceAlcohol` because they are all about alcohol use. 
If we left them in, the tree would just memorize the answer.

## Split + Resample

Here, we stratify the train/test split.

```{r}
#| label: train-test-stratification

set.seed(2023)

alcohol_split <- initial_split(
    riskyBehaviors_analysis,
    strata = UsedAlcohol
)

alcohol_train <- training(alcohol_split)
alcohol_test <- testing(alcohol_split)

alcohol_split
```

We use `strata = ` to keep the proportion of `UsedAlcohol` roughly equal in the train and test sets. 
We use the same logic in the Lasso training.