---
title: "K-Nearest Neighbors"
author: "Daniel Soto"
---

In this note, we use K-Nearest Neighbors to predict whether teens will bring weapons to school.

## Setup

We load our usual packages.

```{r}
#| label: libraries
#| output: false

library(tidyverse)
library(tidymodels)
library(MLearnYRBSS)
library(themis)        # <- for step_downsample()
library(kknn)
library(janitor)
```
Then we get our cached data.

```{r}
#| label: cached-data

analysis_split <- readRDS(here("models", "data", "analysis_split.rds"))
analysis_data <- readRDS(here("models", "data", "analysis_data.rds"))
analysis_train <- readRDS(here("models", "data", "analysis_train.rds"))
analysis_test <- readRDS(here("models","data", "analysis_test.rds"))
analysis_folds <- readRDS(here("models","data", "analysis_folds.rds"))
```

### Sanity check: class balance

As usual, we check the class balance.

```{r}
#| label: sanity-check

analysis_train |>
    tabyl(WeaponCarryingSchool) |>
    adorn_pct_formatting(0) |>
    adorn_totals()

analysis_test |>
    tabyl(WeaponCarryingSchool) |>
    adorn_pct_formatting(0) |>
    adorn_totals()
```
Though the 0/1 split is the same in the training and testing samples, we see that it is especially uneven.
KNN is quite sensitive to this — it will just keep predicting the majority class.
We will fix this in the recipe with `step_downsample()`.

## Workflow

### Recipe

Our recipe is as follows:

```{r}
#| label: recipe

weapon_recipe <-
    recipe(WeaponCarryingSchool ~ ., data = analysis_train) |>
    step_downsample(WeaponCarryingSchool, under_ratio = 1) |>
    step_zv(all_predictors()) |>
    step_normalize(all_numeric_predictors()) |>
    step_impute_mode(all_nominal_predictors()) |>
    step_impute_mean(all_numeric_predictors()) |>
    step_dummy(all_nominal_predictors())

weapon_recipe
```
Here, 

- `step_downsample()` — randomly drops rows from the majority class to fix imbalance. (Critical for KNN.)
- `step_zv()` — drops zero-variance predictors (no info, just noise).
- `step_normalize()` — mandatory for KNN. Without it, predictors with bigger units dominate the distance.
- `step_impute_*()` — KNN can’t compute distance with `NA`s.
- `step_dummy()` — `kknn` needs numeric predictors only.

### Spec — `kknn`

```{r}
#| label: kknn-spec

kknn_spec <-
    nearest_neighbor(
        neighbors = tune()    # ⬅️ we'll tune K
    ) |>
    set_mode("classification") |>
    set_engine("kknn")

kknn_spec
```
### The workflow

```{r}
#| label: workflow

kknn_workflow <-
    workflow() |>
    add_recipe(weapon_recipe) |>
    add_model(kknn_spec)

kknn_workflow

```
## Tuning `K`

### Build the tuning grid

We build a tuning grid with odd numbers. 
This is because with binary classification and odd `K`, we can never get a "tie vote."

```{r}
#| label: neighbor-grid

neighbor_grid <- tibble(neighbors = c(1, 3, 5, 7, 11, 15, 21, 31, 51))
neighbor_grid
```

### Run the tuning

KNN tuning is fast — no model is actually trained, just distances are computed.

```{r}
#| label: run-tuning

doParallel::registerDoParallel()

set.seed(2023)

knn_tune <- tune_grid(
    kknn_workflow,
    resamples = analysis_folds,
    grid      = neighbor_grid,
    metrics   = metric_set(roc_auc, accuracy)
)

doParallel::stopImplicitCluster()
```

### Show the best `K`

Top 5 `K` values, ranked by mean ROC AUC across the 5 folds.

```{r}
#| label: show-best-k

show_best(knn_tune, metric = "roc_auc")
```

### Visualize the tuning results

```{r}
#| label: tuning-visualization

autoplot(knn_tune)
```
### Pick the best `K`

We pick one number as our "winning" `K` value:

```{r}
#| label: best-k

best_k <- select_best(knn_tune, metric = "roc_auc")
best_k
```

But we don't commit to this yet. Instead, we check it first.

## Sanity-Check Before Committing