---
title: "Lasso Regression"
editor_options: 
  chunk_output_type: console
---


Lasso regression is a statistical model that combines linear/logistic regression with L1 regularization to perform both variable selection and regularization. The term "Lasso" stands for "Least Absolute Shrinkage and Selection Operator." This method is particularly useful when dealing with datasets that have many predictors, as it helps to:
- Reduce overfitting by penalizing large coefficients
- Perform automatic feature selection by shrinking some coefficients to exactly zero
- Handle multicollinearity by selecting only one variable from a group of highly correlated predictors

In this analysis, we'll use Lasso regression to predict weapon carrying behavior in schools, demonstrating how this method can help identify the most important predictors while maintaining model interpretability.

## Setting Up the Environment

First, we need to load the necessary packages for our analysis. We'll use `tidymodels` for modeling, `tidyverse` for data manipulation, and `here` for consistent file paths.

```{r}
#| label: packages
#| output: false

library(here)
library(tidymodels)
library(tidyverse)
```

## Loading the Data

We'll work with pre-processed data sets that have been split into training and test sets, along with cross-validation folds. These files are stored in the `processed_data` directory.

```{r}
#| label: load-data

analysis_data <- readRDS(here("models", "data", "analysis_data.rds"))
analysis_train <- readRDS(here("models", "data", "analysis_train.rds"))
analysis_test <- readRDS(here("models","data", "analysis_test.rds"))
analysis_folds <- readRDS(here("models", "data", "analysis_folds.rds"))
```

## Data Preprocessing

Before fitting our model, we need to preprocess the data. We'll create a recipe that:
- Imputes missing values in categorical variables using the mode
- Imputes missing values in numeric variables using the mean
- Removes predictors with zero variance
- Removes highly correlated predictors (correlation threshold = 0.7)
- Creates dummy variables for categorical predictors

```{r}
#| label: recipe

weapon_carry_recipe <- 
  recipe(formula = WeaponCarryingSchool ~ ., data = analysis_train) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_impute_mean(all_numeric_predictors()) |>
  step_zv(all_predictors()) |> 
  step_corr(all_numeric_predictors(), threshold = 0.7) %>% 
  step_dummy(all_nominal_predictors())

weapon_carry_recipe
```

Let's apply our recipe to transform the data according to these preprocessing steps.

```{r}
#| label: bake-recipe

weapon_carry_recipe %>% 
  prep() %>% 
  bake(new_data = analysis_data) 
```

## Model Specification

We'll use a logistic regression model with Lasso regularization. The Lasso (Least Absolute Shrinkage and Selection Operator) helps with feature selection by penalizing the absolute size of coefficients. We set `mixture = 1` to specify a pure Lasso model, and we'll tune the penalty parameter to find the optimal level of regularization.

```{r}
#| label: model-spec

weapon_carry_spec <-
  logistic_reg(penalty = tune(), 
               mixture = 1) |> 
  set_engine('glmnet')

weapon_carry_spec
```

## Creating the Workflow

We'll combine our recipe and model specification into a single workflow. This ensures that all preprocessing steps are properly applied during both training and prediction.

```{r}
#| label: model-workflow

weapon_carry_workflow <-
  workflow() |>
  add_recipe(weapon_carry_recipe) |>
  add_model(weapon_carry_spec)

weapon_carry_workflow
```

## Model Tuning

To find the optimal penalty value, we'll create a grid of potential values to test. We'll use 50 different penalty values, evenly spaced on a logarithmic scale.

```{r}
#| label: create-grid-regular

lambda_grid <- grid_regular(penalty(), levels = 50)
lambda_grid
```

Now, we'll perform cross-validation to find the best penalty value. This process is time-consuming, so we'll save the results for future use.

```{r}
#| label: lasso-tune

set.seed(2023)

lasso_tune <- 
  tune_grid(
  object = weapon_carry_workflow, 
  resamples = analysis_folds,
  grid = lambda_grid, 
  control = control_resamples(event_level = "second")
)
```

```{r}
#| label: save-lasso-tune-rds
#| echo: false

saveRDS(lasso_tune, here("model_outputs", "lasso_tune.rds"))
lasso_tune <- readRDS(here("models", "model_outputs", "lasso_tune.rds"))
```

Let's examine the performance metrics for different penalty values.

```{r}
#| label: collect-metrics

lasso_tune %>% 
  collect_metrics()
```

We can visualize how the model's performance changes with different penalty values.

```{r}
#| label: autoplot

autoplot(lasso_tune)
```

## Selecting the Best Model

We'll select the best model based on the ROC AUC metric, which measures the model's ability to distinguish between classes.

```{r}
#| label: best-lasso-model

best <- lasso_tune |> 
  select_best(metric ="roc_auc")

best
```

Now we'll create our final workflow with the best penalty value.

```{r}
#| label: final-workflow

final_wf <- finalize_workflow(weapon_carry_workflow, best)

final_wf
```

## Fitting the Final Model

We'll fit our final model on the training data. This process is also time-consuming, so we'll save the results.

```{r}
#| label: fit-final-model

weapon_fit <- 
  fit(final_wf, data = analysis_train)

weapon_fit
```

```{r}
#| label: save-weapon-fit-rds
#| echo: false

saveRDS(weapon_fit, here("model_outputs", "weapon_fit.rds"))
weapon_fit <- readRDS(here("models","model_outputs", "weapon_fit.rds"))
```

## Model Evaluation

Let's examine the model's predictions on the training data.

```{r}
#| label: model-eval

weapon_pred <- 
  augment(weapon_fit, analysis_train) |> 
  select(WeaponCarryingSchool, .pred_class, .pred_1, .pred_0)

weapon_pred
```

We can visualize the model's performance using an ROC curve.

```{r}
#| label: roc-curve-training

roc_plot_training <- 
  weapon_pred |> 
  roc_curve(truth = WeaponCarryingSchool, .pred_1, event_level = "second") |> 
  autoplot()

roc_plot_training 
```

Let's look at the model coefficients to understand which predictors are most important.

```{r}
#| label: model-coefficients

weapon_fit |> 
  extract_fit_parsnip() |> 
  tidy()
```

## Cross-Validation Results

We'll fit the model on each cross-validation fold to get a more robust estimate of its performance.

```{r}
#| label: weapon-fit-resamples

weapon_fit_resamples <- 
  fit_resamples(final_wf, resamples = analysis_folds)

weapon_fit_resamples
```

```{r}
saveRDS(weapon_fit_resamples, here("model_outputs", "weapon_fit_resamples.rds"))
```

```{r}
weapon_fit_resamples <- readRDS(here("models","model_outputs", "weapon_fit_resamples.rds"))
```

Let's examine the cross-validation metrics.

```{r}
#| label: cv-results

collect_metrics(weapon_fit_resamples)
```

## Variable Importance

Finally, let's create a variable importance plot to identify the most influential predictors in our model.

```{r}
#| label: vip

library(vip)

weapon_fit |> 
  extract_fit_engine() |> 
  vip() 
```

## Results and Interpretation

The variable-importance ranking suggests that immediate instability and exposure to violence or victimization are the strongest predictors of adolescent weapon-carrying behavior. Homelessness stands out by a substantial margin as the most influential variable, followed by being attacked in the neighborhood and experiencing parental physical abuse. Parental monitoring also contributes meaningfully, while bullying and sexual abuse by an older person form a second tier of important predictors.

School-related factors, including unfair discipline and school connectedness, appear considerably less influential in the model, while parental substance use and especially parental incarceration contribute relatively little once the other predictors are taken into account.

One important qualification: variable importance measures predictive contribution, not direction or causation. This plot alone does not tell us whether a predictor increases or decreases the likelihood of weapon carrying. For example, the importance of parental monitoring could reflect either low monitoring as a risk factor or high monitoring as a protective factor, depending on how the variable was coded and how the model behaves.

Thus, weapon-carrying behavior was predicted most strongly by homelessness and direct exposure to violence or abuse. Homelessness was substantially more important than any other predictor, followed by neighborhood victimization and parental physical abuse. Parental monitoring, bullying, and sexual abuse also contributed meaningfully, whereas school-related factors, parental substance use, and parental incarceration had comparatively lower predictive importance. However, variable importance reflects predictive contribution and should not be interpreted as indicating either the direction of association or causation.
