---
title: "Random Forests"
author: "Daniel Soto"
---

## Setup

```{r}
library(tidyverse)
library(tidymodels)
library(MLearnYRBSS)
library(janitor)
library(skimr)
library(vip)
```

### Load cached data

```{r}
analysis_data <- readRDS(here("models", "data", "analysis_data.rds"))
analysis_train <- readRDS(here("models", "data", "analysis_train.rds"))
analysis_test <- readRDS(here("models","data", "analysis_test.rds"))
analysis_folds <-  readRDS(here("models","data", "analysis_folds.rds"))
```

### Sanity check: class balance

```{r}
#| label: sanity-check

analysis_train |>
    tabyl(WeaponCarryingSchool) |>
    adorn_pct_formatting(0) |>
    adorn_totals()

analysis_test |>
    tabyl(WeaponCarryingSchool) |>
    adorn_pct_formatting(0) |>
    adorn_totals()
```
Both sets have the same proportions, as before.

## Workflow

### Recipe (Preprocessing)

```{r}
hallucinogenic_recipe <-
    recipe(EverHallucinogenicDrugs ~ ., data = hallucinogenic_train) |>
    step_impute_mode(all_nominal_predictors()) |>
    step_impute_mean(all_numeric_predictors()) |>
    step_dummy(all_nominal_predictors())

hallucinogenic_recipe
```

### Random Forest Spec - `ranger`

```{r}
ranger_spec <-
    rand_forest(
        mtry  = tune(),     # ⬅️ # of predictors sampled at each split
        min_n = tune(),     # ⬅️ minimum rows to keep splitting
        trees = 100         # forest size (fixed, not tuned)
    ) |>
    set_mode("classification") |>
    set_engine("ranger",
               importance = "permutation")  # ⬅️ for vip() later

ranger_spec
```
### Workflow

```{r}
ranger_workflow <-
    workflow() |>
    add_recipe(hallucinogenic_recipe) |>
    add_model(ranger_spec)

ranger_workflow
```
### Tuning

```{r}
doParallel::registerDoParallel()

set.seed(46257)

ranger_tune <-
    tune_grid(
        ranger_workflow,
        resamples = cv_hallucinogenic,
        grid = 11    # ⬅️ tidymodels picks 11 sensible combos for us
    )

doParallel::stopImplicitCluster()
```

### Show the best candidates

```{r}
show_best(ranger_tune, metric = "roc_auc")

autoplot(ranger_tune)
```

### Pick the best combo 

```{r}
best_rf <- select_best(ranger_tune, metric = "roc_auc")

best_rf
```

### Ranking the performance of our chosen combo

```{r}
ranger_workflow |>
    finalize_workflow(best_rf) |>
    fit_resamples(resamples = cv_hallucinogenic) |>
    collect_metrics()
```
### Looks good - locking in

```{r}
final_wf <- finalize_workflow(ranger_workflow, best_rf)

final_wf
```
### Evaluate the last test set

```{r}
hallucinogenic_last_fit <-
    last_fit(
        final_wf,
        split   = hallucinogenic_split,
        metrics = metric_set(accuracy, kap, roc_auc, sensitivity, specificity)
    )

hallucinogenic_last_fit
collect_metrics(hallucinogenic_last_fit)
```

### Test-set predictions

```{r}
predictions_testing <-
    hallucinogenic_last_fit |>
    collect_predictions() |>
    select(-.config, -.row)

predictions_testing
```

### Confusion matrix

```{r}
predictions_testing |>
   conf_mat(truth = EverHallucinogenicDrugs, estimate = .pred_class) |>
    autoplot(type = "heatmap")
```

### ROC curve on the test set

```{r}
predictions_testing |>
    roc_curve(truth = EverHallucinogenicDrugs, .pred_1, event_level = "second") |>
    autoplot()
```

### Variable Importance

```{r}
hallucinogenic_last_fit |>
    extract_fit_parsnip() |>
    vi() |>
    slice_max(Importance, n = 10) |>
    ggplot(aes(Importance, fct_reorder(Variable, Importance))) +
    geom_col(fill = "forestgreen") +
    labs(y = NULL, title = "Top 10 predictors of Usedhallucinogenic")
```