---
title: "Decision Trees"
author: "Daniel Soto"
---

In this note, we check whether a decision tree can predict whether a teen carries a weapon to school.

## Setup

Like before we load the packages necessary for the analysis.

```{r}
#| label: packages
#| output: false

library(tidyverse)
library(tidymodels)
library(MLearnYRBSS)
library(janitor)
library(skimr)
library(vip)
library(here)
library(rpart.plot)
```

And we load the data from before!

```{r}
#| label: load-data

analysis_split <- readRDS(here("models", "data", "analysis_split.rds"))
analysis_data <- readRDS(here("models", "data", "analysis_data.rds"))
analysis_train <- readRDS(here("models", "data", "analysis_train.rds"))
analysis_test <- readRDS(here("models","data", "analysis_test.rds"))
```

Except this time, we use define `analysis_folds` with `v = 10` folds, meaning every candidate tree we try will be evaluated on all 10 — the test set stays untouched until the very end. 
Further, explicitly specify `strata = WeaponCarryingSchool`:

```{r}
#| label: cv-folds

analysis_folds <- vfold_cv(analysis_train, v = 10, strata = WeaponCarryingSchool)
analysis_folds
```

### Sanity check: class balance

Before we continue, we need to check the class balances between the train and test sets:

```{r}
#| label: class-balance-check

analysis_train |>
    tabyl(WeaponCarryingSchool) |>
    adorn_pct_formatting(0) |>
    adorn_totals()

analysis_test |>
    tabyl(WeaponCarryingSchool) |>
    adorn_pct_formatting(0) |>
    adorn_totals()
```
Both sets show similar proportions, meaning `strata = WeaponCarryingSchool` did its job when we defined the train/test split previously.
However, we note that there is a significant difference in proportion between those who carry weapons (0) and those who do not (1): we will need to address this in the specification section.

## Workflow

### Recipe (preprocessing)

Trees don't need fancy preprocessing: no `step_normalize()`, as trees only care about *order*, not scale; no `step_dummy()`, as `rpart` handles factor predictors natively; and imputation is still useful, since `rpart` can't split on `NA`. 

```{r}
#| label: recipe

weapon_carry_recipe <- 
  recipe(formula = WeaponCarryingSchool ~ ., data = analysis_train) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_impute_mean(all_numeric_predictors()) |>
  step_zv(all_predictors())

weapon_carry_recipe
```

### Specification

We do the specification as follows.

As a result of the proportions of weapon-carrying being very unequal, when we set the `rpart` engine, we add the parameter `parms = list(prior = c(0.5, 0.5))`.
This tells the `rpart` engine to treat each class equally, despite the proportional class imbalance.

```{r}
#| label: cart-spec

cart_spec <-
    decision_tree(
        cost_complexity = tune(),
        tree_depth = tune(),
        min_n = tune()
    ) |>
    set_engine(
        "rpart",
        parms = list(prior = c(0.5, 0.5))
    ) |>
    set_mode("classification")

cart_spec
```

### Workflow

Bundle it! This is the recipe and specification, neatly packaged.

```{r}
#| label: workflow

cart_workflow <-
    workflow() |>
    add_recipe(weapon_carry_recipe) |>
    add_model(cart_spec)

cart_workflow
```

## Tree tuning

### Tuning grid

We first set up the tuning grid. 
Note that `levels` can be set to `3` for faster turnaround.

```{r}
#| label: tuning-grid

tree_grid <- grid_space_filling(
    cost_complexity(range = c(-6, -1)),
    tree_depth(c(2, 10)),
    min_n(c(5, 150)),
    size = 60
)

tree_grid
```

### Run the tuning

Note for those following along that this part can be slow, as there are:

- 4 × 4 × 4 = **64 candidates**
- × 10 folds = **640 model fits**

We’ll use `doParallel` to speed it up.
Also, we use ROC AUC as the tuning metric — robust to class imbalance, and the same one we used for Lasso. 

```{r}
#| label: run-tuning

doParallel::registerDoParallel()

cart_tune <-
    cart_workflow |>
    tune_grid(
        resamples = analysis_folds,
        grid = tree_grid,
        metrics = metric_set(roc_auc),
        control = control_grid(save_pred = TRUE)
    )

doParallel::stopImplicitCluster()
```

### Show the best candidates

The top 5 hyperparameter combos, ranked by mean ROC AUC across the 10 folds.

```{r}
show_best(cart_tune, metric = "roc_auc")
```

### Visualize the tuning results

In this visualization, each panel shows how the metric changes as one knob varies — colored by the others. 
We look for the sweet spot.

```{r}
autoplot(cart_tune)
```

### Picking the best combination

We select one row as the winner candidate:

```{r}
best_cart <- select_best(cart_tune, metric = "roc_auc")

best_cart
```

But we don't commit yet: we want to check how this performs first.

## How does our chosen candidate actually do?

Here, we get an honest, CV-estimated performance for our single chosen combo.

```{r}
cart_workflow |>
    finalize_workflow(best_cart) |>
    fit_resamples(resamples = analysis_folds) |>
    collect_metrics()
```

Looks good, time to lock it in.

```{r}
cart_final_wf <- finalize_workflow(cart_workflow, best_cart)

cart_final_wf
```

`finalize_workflow()` swaps the `tune()` placeholders for the actual best values. 
Now we’re committed.

## Evaluate on the Test Set

We use `last_fit()` to fit **all** training data to predit on the **test** set.
It computes metrics once, and honestly, with no accidental cheating possible.

```{r}
#| label: last-fit

cart_last_fit <-
    last_fit(
        cart_final_wf,
        split   = analysis_split,
        metrics = metric_set(accuracy, kap, roc_auc, sensitivity, specificity)
    )

cart_last_fit
```

### Test-set metrics

Finally, we obtain our metrics:

```{r}
#| label: test-set-metrics

collect_metrics(cart_last_fit)
```

Comparing these to the CV metrics from the sanity-check section, we see they are in about the same ballpark.
Thus, the model generalizes.

### Test-set predictions

```{r}
#| label: test-set-predictions

tree_pred <-
    cart_last_fit |>
    collect_predictions() |>
    select(WeaponCarryingSchool, .pred_class, .pred_1, .pred_0)

tree_pred
```

### Confusion Matrix

```{r}
#| label: confusion-matrix

tree_pred |>
    conf_mat(truth = WeaponCarryingSchool, estimate = .pred_class) |>
    autoplot(type = "heatmap")
```

### ROC curve on the test set

```{r}
#| label: ROC-curve

tree_pred |>
    roc_curve(truth = WeaponCarryingSchool, .pred_1, event_level = "second") |>
    autoplot()
```

We see the ROC curve hugs the top-left corner, which is excellent and means the model is not randomly guessing.

## Visualizing the final tree

This is the superpower of trees: it is a model your collaborator (or PI, or grandma) can read without a stats degree.

```{r}
#| label: tree-visualization

cart_last_fit |>
    extract_fit_engine() |>
    rpart.plot(roundint = FALSE, type = 4, extra = 104)
```
## Variable importance with `vip`

Each bar in the variable importance plot (VIP) indicates how much that predictor reduced impurity across all splits.
Predictors not shown were never useful enough to make the tree.
This makes for a free feature ranking for your manuscript.

```{r}
#| label: vip

cart_last_fit |>
    extract_fit_parsnip() |>
    vip(num_features = 15)
```